---
title: "Cardinal tutorial"
author: "Kylie Ariel Bemis"
date: "10/10/2019"
output: beamer_presentation
---
  
  ```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```

## Slides and code

The R Markdown for this tutorial is available on Github at:

https://github.com/kuwisdelu/asilomar2019

Use R or RStudio to run the document (you may need to update some file paths).

## Installation

First make sure `BiocManager` is installed.

Then use `BiocManager` to install `Cardinal`.

```{r install, eval=FALSE}
if ( !requireNamespace("BiocManager") )
  install.packages("BiocManager")

BiocManager::install("Cardinal")
```

For the datasets used in the statistical analysis examples, we'll also install `CardinalWorkflows`.

```{r install-workflows, eval=FALSE}
BiocManager::install("CardinalWorkflows")
```

## Getting started

Load the Cardinal package with `library()`.

```{r library, message=FALSE}
library(Cardinal)

register(SerialParam())
```

*Note: Cardinal supports parallelization via the `BiocParallel` package. The code above registers a serial (i.e., non-parallel) backend for this session.*



# Cardinal basics

## Example data

To demonstrate reading data and visualization, we'll use [example imzML files from imzml.org](https://ms-imaging.org/wp/imzml/example-files-test/).

Specifically, we are going to reproduce the first image from the imzml.org website, which comes from the paper *"Histology by mass spectrometry: label-free tissue characterization obtained from high-accuracy bioanalytical imaging"* by Rompp et al.

The data is available from https://www.ebi.ac.uk/pride/archive/projects/PXD001283.

## Importing data

We can use `readMSIData()` to import data from an imzML file.

\small
```{r mouse}
file <- file.path("~/Documents/Datasets/PRIDE/PXD001283",
                  "HR2MSI mouse urinary bladder S096.imzML")

mouse <- readMSIData(file, mass.range=c(400, 900),
                     resolution=0.01, units="mz")
```
\normalsize

Providing the mass range and binning resolution will allow faster importing; otherwise, `Cardinal` will try to infer them from reading the data.

---

```{r mouse-show}
mouse
```

## Accessing pixel metadata

\scriptsize
```{r mouse-pdata}
pixelData(mouse)
```
\normalsize

## Accessing feature metadata

\scriptsize
```{r mouse-fdata}
featureData(mouse)
```
\normalsize

## Accessing the spectra matrix

\scriptsize
```{r mouse-spectra}
spectra(mouse)
```
\normalsize

## Visualize mass spectra

```{r mouse-plot-1, fig.height=4, fig.width=9}
plot(mouse, coord=list(x=150, y=100))
```

## Visualize ion images

```{r mouse-image-1, fig.height=3, fig.width=6}
image(mouse, mz=798.54, plusminus=0.01)
```

## Visualize ion images (w/ enhancement)

```{r mouse-image-2, fig.height=3, fig.width=6}
image(mouse, mz=798.54, plusminus=0.01,
      contrast.enhance="suppress",
      smooth.image="gaussian")
```

## Visualize ion images (superposed)

```{r mouse-image-3, fig.height=3, fig.width=6}
image(mouse, mz=c(743.5483, 798.5419, 741.5307),
      plusminus=0.01, superpose=TRUE,
      contrast.enhance="suppress",
      normalize.image="linear",
      col=c("red", "green", "blue"))
```



# Segmentation

## Pig fetus section

We will use some pre-processed datasets and pre-calculated analysis results from the `CardinalWorkflows` package.

For segmentation, we will use data from a pig fetus cross-section.


```{r pig-1, message=FALSE}
data(pig206, package="CardinalWorkflows")

pig206_peaks <- as(pig206.peaks, "MSImagingExperiment")
```

\scriptsize
*Note: The datasets provided in `CardinalWorkflows` are stored in an older format, so we need to coerce them to an `MSImagingExperiment` using `as()`.*
\normalsize

## Pig fetus m/z 888

```{r pig-image, fig.height=3, fig.width=6}
image(pig206_peaks, mz=888)
```

## Spatial shrunken centroids

We can use `spatialShrunkenCentroids()` to perform spatial shrunken centroids clustering.

This method requires three parameters:

- `r` = smoothing radius

- `k` = initial number of segments

- `s` = sparsity parameter

The method performs automated feature selection controlled by the sparsity parameter `s`.

\small
```{r pig-ssc, eval=FALSE}
set.seed(2019)
pig206_sscg <- spatialShrunkenCentroids(pig206_peaks,
                                        r=1, k=10,
                                        s=c(0,3,6,9),
                                        method="gaussian")
```
\normalsize

## Loading pre-calculated segmentation results

To save time, we will load some pre-calculated segmentation results.

We'll load results for two types of spatial smoothing (`method="gaussian"` and `method="adaptive").

\small
```{r pig-ssc-load}
data(pig206_analyses, package="CardinalWorkflows")
pig206_sscg <- as(pig206.sscg, "SpatialShrunkenCentroids2")
pig206_ssca <- as(pig206.ssca, "SpatialShrunkenCentroids2")
```
\normalsize

\scriptsize
*Note: The datasets provided in `CardinalWorkflows` are stored in an older format, so we need to coerce them to an `MSImagingExperiment` using `as()`.*
\normalsize

---

\scriptsize
```{r pig-sscg}
summary(pig206_sscg)
```
\normalsize

---

\scriptsize
```{r pig-ssca}
summary(pig206_ssca)
```
\normalsize

## Probability of segment membership (Gaussian smoothing)

```{r pig-sscg-prob, fig.height=3, fig.width=6}
image(pig206_sscg, model=list(r=2, k=20, s=9),
      values="probability")
```

## Probability of segment membership (adaptive smoothing)

```{r pig-ssca-prob, fig.height=3, fig.width=6}
image(pig206_ssca, model=list(r=1, k=20, s=9),
      values="probability")
```

## Brain, liver, and heart segments

```{r pig-sscg-cols, fig.height=3, fig.width=6}
image(pig206_sscg, model=list(r=2, k=20, s=9),
      values="probability", column=c(3,5,6))
```

## Mean spectra for brain, liver, and heart

```{r pig-sscg-centers, fig.height=3, fig.width=6}
plot(pig206_sscg, model=list(r=2, k=20, s=9),
     values="centers", column=c(3,5,6), lwd=2)
```

## Statistics for brain, liver, and heart

```{r pig-sscg-stats, fig.height=3, fig.width=6}
plot(pig206_sscg, model=list(r=2, k=20, s=9),
     values="statistic", column=c(3,5,6), lwd=2)
```

## Most distinctive mass features for heart

\small
```{r pig-sscg-top}
topFeatures(pig206_sscg, model=list(r=2, k=20, s=9), class==6)
```
\normalsize

## Most distinctive mass features for heart

```{r pig-image-regions, fig.height=3, fig.width=9}
image(pig206_peaks, mz=c(187.3, 186.4, 179.4),
      normalize.image="linear", layout=c(1,3))
```




# Classification

## Renal cell carcinoma (RCC) dataset

For classification, we will look at a dataset with 8 matched pairs of healthy tissue and tumor.

```{r rcc-1}
data(rcc, package="CardinalWorkflows")

rcc_binned <- as(rcc.small, "MSImagingExperiment")
```

\scriptsize
*Note: The datasets provided in `CardinalWorkflows` are stored in an older format, so we need to coerce them to an `MSImagingExperiment` using `as()`.*
\normalsize

## Diagnosis

```{r rcc-diagnosis, fig.height=4, fig.width=9}
image(rcc_binned, diagnosis ~ x * y,
      layout=c(2,4), key=TRUE)
```

## RCC m/z 810

```{r rcc-image, fig.height=4, fig.width=9}
image(rcc_binned, mz=810, layout=c(2,4),
      contrast.enhance="histogram",
      normalize.image="linear")
```

## Cross-validation

We use `crossValidate()` to perform cross-validation with spatial shrunken centroids classification.

Each run (with a cancer/normal matched pair) is a separate CV fold.

```{r rcc-cv, eval=FALSE}
rcc_cv <- crossValidate(rcc_binned,
                        .y=rcc_binned$diagnosis,
                        .fun=spatialShrunkenCentroids,
                        r=1, s=0:9,
                        .fold=run(rcc_binned))
```

## Loading pre-calculated CV results

To save time, we will load some pre-calculated CV results.

```{r rcc-cv-load}
data(rcc_analyses, package="CardinalWorkflows")

rcc_cv <- as(rcc.cv.sscg, "CrossValidated2")
rcc_cv$.response <- rcc_binned$diagnosis
```

\scriptsize
*Note: The CV results provided in `CardinalWorkflows` are stored in an older format, so we need to coerce them to the newer version of the class using `as()`.*
\normalsize

---

\tiny
```{r rcc-cv-summary}
summary(rcc_cv)
```
\normalsize

---

Which is the most accurate model?

\scriptsize
```{r rcc-cv-best}
i <- which.max(summary(rcc_cv)$Accuracy)
summary(rcc_cv)[i,]
```
\normalsize

---

Fit the model with the best parameters to the full dataset.

\scriptsize
```{r rcc-sscg}
rcc_sscg <- spatialShrunkenCentroids(rcc_binned,
                                   y=rcc_binned$diagnosis,
                                   r=3, s=20)
summary(rcc_sscg)
```
\normalsize

## Predicted probability of cancer/normal

```{r rcc-sscg-prob, fig.height=4, fig.width=9}
image(rcc_sscg, values="probability", layout=c(2,4))
```

## Mass features over/under-expressed in each class

```{r rcc-sscg-stat, fig.height=4, fig.width=9}
plot(rcc_sscg, values="statistic", lwd=2)
```

## Mass features associated with cancer

```{r rcc-sscg-top-cancer}
topFeatures(rcc_sscg, class=="cancer")
```

## m/z 886 associated with cancer

```{r rcc-cancer-image, fig.height=4, fig.width=9}
image(rcc_binned, mz=886, layout=c(2,4),
      contrast.enhance="histogram",
      normalize.image="linear")
```

## Mass features associated with normal

```{r rcc-sscg-top-normal}
topFeatures(rcc_sscg, class=="normal")
```

## m/z 215 associated with normal

```{r rcc-normal-image, fig.height=4, fig.width=9}
image(rcc_binned, mz=215, layout=c(2,4),
      contrast.enhance="histogram",
      normalize.image="linear")
```




# Statistical testing

## Statistical testing

Suppose we want to test which mass features are differentially abundant between cancer and normal tissue?

We have N = 8 subjects per condition.

Two approaches we might take using `Cardinal`:

- Average all pixels in each tissue sample and compare means

- Segment each tissue and compare distinctive segments

Let's explore both.

## Filter mass features by intensity thresholding

\small
```{r rcc-2}
rcc_filt <- rcc_binned %>%
  mzFilter(thresh.max=0.25) %>%
  process()

rcc_filt
```
\normalsize

## Means-based testing

We need to create a variable for each (subject x condition) combination.

Then we can use `meansTest()` to calculate the mean intensities for each group and fit linear models for each mass feature.

```{r mtest-fit}
rcc_filt$sample <- interaction(run(rcc_filt),
                               rcc_filt$diagnosis)

rcc_mtest <- meansTest(rcc_filt, ~ diagnosis,
                       groups=rcc_filt$sample)
```

---

\tiny
```{r mtest-summary}
summary(rcc_mtest)
```
\normalsize

## Significant mass features

```{r mtest-top}
topFeatures(rcc_mtest, p.adjust="fdr", AdjP < 0.05)
```

---

```{r mtest-plot}
plot(rcc_mtest, model=c(2,19,22,23))
```

---

```{r mtest-image, fig.height=4, fig.width=9}
image(rcc_mtest, model=22, layout=c(2,4))
```

## Segmentation-based testing

To represent potentialy heterogeneity within the tissue, we can segment each tissue first and then compare the most mean intensities for the segments.

First we use `spatialDGMM()` to fit segmentations for each mass feature, and then `segmentationTest()` fits linear models to the segments.

```{r stest-fit}
rcc_dgmm <- spatialDGMM(rcc_filt,
                        r=1, k=2,
                        groups=rcc_filt$sample)

rcc_stest <- segmentationTest(rcc_dgmm, ~ diagnosis)
```

---

\tiny
```{r stest-summary}
summary(rcc_stest)
```
\normalsize

## Significant mass features

```{r stest-top}
topFeatures(rcc_stest, p.adjust="fdr", AdjP < 0.05)
```

---

```{r stest-plot}
plot(rcc_stest, model=c(2,19,22,23))
```

---

```{r stest-image, fig.height=4, fig.width=9}
image(rcc_stest, model=22, layout=c(2,4))
```



# Interactive visualization

## Interactive visualization

Shiny interface for Cardinal (under development) on Github...

```{r eval=FALSE}
if (!requireNamespace("remotes", quietly = TRUE))
    install.packages("remotes")

BiocManager::install("kuwisdelu/CardinalVis")

library(CardinalVis)

pig206 <- as(pig206, "MSImagingExperiment")

msiVis(pig206)
```

# Questions

